The Social Conditioning of Fear

### 2023-12-06 scripts/manuscript.qmd
background

This is an exploratory project for a dissertation chapter idea that one of the co-authors wants to pursue. With descriptive statistics, we are interested in investigating the inequalities in terms of fear of walking alone at night among US adult population and mapping out the patterns of disparities along the axes of race, gender, sexual orientation, citizenship status, and education.

@@

Especially, the role of gender in fear of crime has been well documented in victimization research [@cops2011; @snedker2006; @schnitzer2010; @stanko1993]. Early scholarship emphasized that women tend to report more fear than men  [@stanko1990], with some initial research investigating the seeming disjuncture between women's fear and their actual experiences of victimization, as women were often found to report high levels of fear for crimes that they were at relatively little risk of experiencing [@hough1983]. Subsequent studies suggested that women's fear should be understood in the context of the many potential threats that women confront, including their heightened risk of sexual assault [@stanko1993]. Therefore, fear at night is gradually perceived as a manifestation of gender-based violence with sexual assault implications, though acknowledging that sexual assault is merely one among many potential forms of crime.

>>>>

@@

Especially, the role of gender in fear of crime has been well documented in victimization research [@cops2011; @snedker2006; @schnitzer2010; @stanko1993]. Early scholarship emphasized that women tend to report more fear than men  [@stanko1990], with some initial research investigating the seeming disjuncture between women's fear and their actual experiences of victimization, as women were often found to report high levels of fear for crimes that they were at relatively little risk of experiencing [@hough1983]. Subsequent studies suggested that women's fear should be understood in the context of the many potential threats that women confront, including their heightened risk of sexual assault [@stanko1993]. Therefore, fear at night is gradually perceived as a manifestation of gender-based violence with sexual assault implications, though acknowledging that sexual assault is merely one among many potential forms of crime. However, the more insidious and everyday forms of harassment are frequently not captured by common survey instruments--they regularly privilege the more obvious or extreme forms of victimization and fail to account for the routine forms of violence that women confront [@moran2003]. We focus our attention on this less intrusive but omnipresent form of everyday violence and examine the unequal effects it has on different populations. We ask: what groups of people are more likely to have experienced fear of walking alone at night in their neighborhood?

Historically, despite all the aforementioned contexts and individual-level characteristics being identified as salient to individuals’ self-reported fear at night, they were usually studied in isolation. Notably, scholars have tended to focus either on individual-level characteristics, such as race and sexual orientation, or on physically defined contexts like schools and neighborhoods and social contexts like networks, families, with little consideration of the joint contributions of multiple contexts and multilevel effects simultaneously. Instead, we hypothesize that, similar to "the social determinants of health" [@krieger2001], there exist the social determinants of fear--intersectional social positionalities impact the likelihood of individuals's sense of threat from social surroundings, i.e., fear at night. This is an exploratory project for a dissertation chapter idea that one of the co-authors wants to pursue. With descriptive statistics, we are interested in investigating the inequalities in terms of fear of walking alone at night among US adult population and mapping out the patterns of disparities along the axes of race, gender, sexual orientation, citizenship status, and education.
